Data Analysis Methods for the Detection of the Epoch of Reionization

### 2017-11-22 PAPER.tex
-Set aside section for Danny/Matt in signal loss

The relationship between the input and output power spectra, $P_{in}$ and $P_{out}$, can be thought of as a transfer function which, for a sampling of $P_{in}$ and $P_{out}$ provides a mapping from an input power spectrum distribution into an output distribution. By viewing data through this signal loss lens, we may then ask the question ``what input power spectrum distribution could this (signal-loss affected) data come from?" In the next section, we describe the shape of our signal loss transfer function and detail the computation used to translate our power spectrum result into one viewed through a signal loss lens.

@@

@@

The solid black diagonal line in Figure \ref{fig:sigloss_transfercurve} shows unity-transfer, which we expect to occur for the unweighted case ($\textbf{R} = \textbf{I}$). Therefore, it is worth thinking about the spread, or scatter, around this black line that we see in the right plot (the colored `heat-map'). One might imagine that there should be one true $P_{out}$ value for every $P_{in}$, implying one well-determined signal loss factor for every data value. Focusing on the unweighted case alone, the scatter observed arises from the fact that the values of the cross-terms involving $\textbf{e}$ and $\textbf{x}$ in Equation \eqref{eq:pout_expand} change for different mock EoR signals, $\textbf{e}$. Since we draw different random $\textbf{e}$'s for every bootstrap, there is a range of $P_{out}$ values that results from the same $P_{in}$. This randomness causes the spread seen in the right plot of Figure \ref{fig:sigloss_transfercurve}, as well as most of the spread in the weighted case (left plot).

There is a second component of scatter that is less intuitive, only applies (at a low level) to the weighted case, and represents intrinsic scatter in signal loss. The origin of this scatter stems from how much the random EoR signal $\textbf{e}$ happens to look like the data $\textbf{x}$. For example, if we are extremely unlucky, one realization of $\textbf{e}$ would have the same shapes, or eigenvectors, as $\textbf{x}$. An empirically-derived covariance would then down-weight these shapes, destroying the entire EoR signal. On the other hand, the less that $\textbf{e}$ looks like $\textbf{x}$, the less signal loss that would result. The intrinsic scatter we can get is not a dominant factor in this case but it is important to correct for the fact that a particular $P_{out}$ value could arise from a range of $P_{in}$ values. We correct for this, as explained later in this section.

@@

To conclude this section, we also show power spectrum results for fringe-rate filtered PAPER-64 data before and after signal loss correction in Figure \ref{fig:ps2_data}, using inverse covariance weighting. The blue dashed line represents the unweighted power spectrum, which is identical in both panels (an important check, as we expect no signal loss for this case). The black and grey points are positive and negative power spectrum values plotted with $2\sigma$ error bars. Prior to signal loss correction, it is clear that the power spectrum is unfeasible because it is well below the theoretical noise level prediction (solid green curve). Post-correction, the power spectrum values blow up to be higher than both the theory and unweighted power spectrum.

Investigating this further, the reason why the post-corrected power spectrum is so high is due to the small number of independent samples in PAPER-64. PAPER-64 only has $\sim$$8$ independent modes after fringe-rate filtering while our power spectrum bandwidth has $21$ channels. Consequently, the eigenvalues of $\hat{\textbf{C}}$ cannot possibly describe real fluctuations beyond the first eight. Hence, there is a substantial down-weighting of many random modes, a large loss that the signal loss transfer function captures and corrects for. We elaborate on this point in the next section, as well as investigate alternate weighting schemes to inverse covariance weighting, with the goal of finding one that balances the aggressiveness of down-weighting contaminants with minimizing the loss of EoR.

@@

@@

With our signal loss formalism established, we now have the capability of experimenting with different weighting options for $\textbf{R}$. Our goal here is to choose a weighting method that successfully down-weights foregrounds and systematics in our data without generating large amounts of signal loss. We have found that the balance between the two is a delicate one and requires a careful understanding and altering of covariance matrices.

We saw in Section \ref{sec:otherweight} how limiting the number of down-weighted eigenmodes (i.e. flattening out part of the eigenspectrum) can help minimize signal loss. We experiment with this idea on PAPER-64 data, dialing the number of modes down-weighted from zero (which is equivalent to identity-weighting, or the unweighted case) to $21$ (which is full inverse covariance weighting of our $21$ channels). The power spectrum results for one $k$ value, both before and after signal loss correction, are shown in Figure \ref{fig:sigloss_modeloop}. We see that the amount of signal loss increases as weighting becomes more aggressive (grey curve). In other words, more random fluctuations are being overfit and subtracted as more modes are down-weighted. We also find that the power spectrum upper limit, post signal loss correction, increases with the number of down-weighted modes (black curve). As discussed previously, we are limited in our power spectrum accuracy by the number of independent samples in our dataset and our errors dramatically increase as we surpass the $\sim8$ independent samples that we have in fringe-rate filtered PAPER-64 data \cc{elaborate?}.

@@

In addition to our thermal noise prediction (green), two additional horizontal lines are shown in Figure \ref{fig:sigloss_modeloop} which denote power spectrum values, post-signal loss correction, for two other regularization schemes. We multiply an identity matrix element-wise to $\hat{\textbf{C}}$ (i.e. inverse variance weighting, shown in red), and also add an arbitrarily chosen level of $\textbf{I}$ to $\hat{\textbf{C}}$ (cyan). We see that all three regularization schemes (black, red, cyan) perform similarly at their best.

We choose to use $\hat{\textbf{C}} = \hat{\textbf{C}} \circ \textbf{I}$, or inverse variance weighting, for the remainder of this paper, which we will denote as $\hat{\textbf{C}}_{eff}$. We choose this weighting scheme due to its simplicity in practice and because its effectiveness is comparable to other weighting schemes that produce minimal signal loss.

@@

Another possible way to regularize $\hat{\textbf{C}}$ is to use information from different ranges of LST. For example, one could calculate $\hat{\textbf{C}}$ with data from LSTs where foregrounds are stronger (earlier or later LSTs than the `foreground-quiet' range used in forming power spectra) --- doing so may yield a better description of the foregrounds that we desire to down-weight. Fundamentally, each of these examples are similar in that they rely on a computation of $\hat{\textbf{C}}$ from data that is similar but not exactly the same as the data that is being down-weighted. Ideally this would be effective in down-weighting shared contaminants yet avoid signal loss from over-fitting modes in the power spectrum dataset itself.

>>>>

The relationship between the input and output power spectra, $P_{in}$ and $P_{out}$, can be thought of as a transfer function which, for a sampling of $P_{in}$ and $P_{out}$ provides a mapping from an input power spectrum distribution into an output distribution. By viewing data through this signal loss lens, we may then ask the question ``what input power spectrum distribution could this (signal-loss affected) data come from?" In the next section, we describe the shape of our signal loss transfer function and detail two different computations used to translate our power spectrum result into one viewed through a signal loss lens. We showcase two methods... \cc{fill in with some broad statement of why we're showing 2 methods but also state that they yield similar results...}

@@

\begin{center}
METHOD \#1
\end{center}

@@

The solid black diagonal line in Figure \ref{fig:sigloss_transfercurve} shows unity-transfer, which we expect to occur for the unweighted case ($\textbf{R} = \textbf{I}$). Therefore, it is worth thinking about the spread, or scatter, around this black line that we see in the right plot (the colored `heat-map'). One might imagine that there should be one true $P_{out}$ value for every $P_{in}$, implying one well-determined signal loss factor for every data value. Focusing on the unweighted case alone, the scatter observed arises from the fact that the values of the cross-terms involving $\textbf{e}$ and $\textbf{x}$ in Equation \eqref{eq:pout_expand} change for different mock EoR signals, $\textbf{e}$. Since we draw different random $\textbf{e}$'s for every bootstrap (for reasons made clear in the next paragraph), there is a range of $P_{out}$ values that results from the same $P_{in}$. This randomness causes the spread seen in the right plot of Figure \ref{fig:sigloss_transfercurve}, as well as most of the spread in the weighted case (left plot).

There is a second component of scatter that is less intuitive, only applies (at a low level) to the weighted case, and represents intrinsic scatter in signal loss. The origin of this scatter stems from how much the random EoR signal $\textbf{e}$ happens to look like the data $\textbf{x}$, a quantity we do not know offhand but one that we would like to correct for. For example, if we are extremely unlucky, one realization of $\textbf{e}$ would have the same shapes, or eigenvectors, as $\textbf{x}$. An empirically-derived covariance would then down-weight these shapes, destroying the entire EoR signal. On the other hand, the less that $\textbf{e}$ looks like $\textbf{x}$, the less signal loss that would result. The intrinsic scatter we can get is not a dominant factor in this case but it is important to correct for the fact that a particular $P_{out}$ value could arise from a range of $P_{in}$ values. We correct for this, as explained later in this section.

@@

To conclude this method, we also show power spectrum results for fringe-rate filtered PAPER-64 data before and after signal loss correction in Figure \ref{fig:ps2_data}, using inverse covariance weighting. The blue dashed line represents the unweighted power spectrum, which is identical in both panels (an important check, as we expect no signal loss for this case). The black and grey points are positive and negative power spectrum values plotted with $2\sigma$ error bars. Prior to signal loss correction, it is clear that the power spectrum is unfeasible because it is well below the theoretical noise level prediction (solid green curve). Post-correction, the power spectrum values blow up to be higher than both the theory and unweighted power spectrum.

Investigating this further, the reason why the post-corrected power spectrum is so high is due to the small number of independent samples in PAPER-64. PAPER-64 only has $\sim$$8$ independent modes after fringe-rate filtering while our power spectrum bandwidth has $21$ channels. Consequently, the eigenvalues of $\hat{\textbf{C}}$ cannot possibly describe real fluctuations beyond the first eight. Hence, there is a substantial down-weighting of other modes, a large loss that the signal loss transfer function captures and corrects for. We elaborate on this point in the next section, as well as investigate alternate weighting schemes to inverse covariance weighting, with the goal of finding one that balances the aggressiveness of down-weighting contaminants with minimizing the loss of EoR.

@@

\begin{center}
METHOD \#2
\end{center}

The second method we describe here... \cc{Danny and Matt's section}

\cc{refer back to Equations \eqref{eq:Pin} and \eqref{eq:sigloss} and explain why the subtraction term is removed}

\cc{also state at the end that the rest of the analysis uses method \#1 instead of \#2...}

@@

With a signal loss formalism established (Method \#1, for the rest of this paper), we now have the capability of experimenting with different weighting options for $\textbf{R}$. Our goal here is to choose a weighting method that successfully down-weights foregrounds and systematics in our data without generating large amounts of signal loss. We have found that the balance between the two is a delicate one and requires a careful understanding and altering of covariance matrices.

We saw in Section \ref{sec:otherweight} how limiting the number of down-weighted eigenmodes (i.e. flattening out part of the eigenspectrum) can help minimize signal loss. We experiment with this idea on PAPER-64 data, dialing the number of modes down-weighted from zero (which is equivalent to identity-weighting, or the unweighted case) to $21$ (which is full inverse covariance weighting of our $21$ channels). The power spectrum results for one $k$ value, both before and after signal loss correction, are shown in Figure \ref{fig:sigloss_modeloop}. We see that the amount of signal loss increases as weighting becomes more aggressive (grey curve). In other words, more `weak' (EoR-dominated) fluctuations are being overfit and subtracted as more modes are down-weighted. We also find that the power spectrum upper limit, post signal loss correction, increases with the number of down-weighted modes (black curve). As discussed previously, we are limited in our power spectrum accuracy by the number of independent samples in our dataset and our errors dramatically increase as we surpass the $\sim8$ independent samples that we have in fringe-rate filtered PAPER-64 data \cc{elaborate?}.

@@

In addition to our thermal noise prediction (green), two additional horizontal lines are shown in Figure \ref{fig:sigloss_modeloop} which denote power spectrum values, post-signal loss correction, for two other regularization schemes. We multiply an identity matrix element-wise to $\hat{\textbf{C}}$ (i.e. inverse variance weighting, shown in red), and also add an arbitrarily chosen level of $\textbf{I}$ to $\hat{\textbf{C}}$ (cyan). We see that all three regularization schemes (black, red, cyan) perform similarly at their best (i.e. when $\sim3$-$4$ eigenmodes are down-weighted in the case of the black curve).

For the remainder of this paper, we choose to use the weighting option of $\hat{\textbf{C}} = \hat{\textbf{C}} \circ \textbf{I}$, or inverse variance weighting, which we will denote as $\hat{\textbf{C}}_{eff}$. We choose this weighting scheme due to its simplicity in practice and because its effectiveness is comparable to other weighting schemes that produce minimal signal loss.

@@

Another possible way to regularize $\hat{\textbf{C}}$ is to use information from different ranges of LST. For example, one could calculate $\hat{\textbf{C}}$ with data from LSTs where foregrounds are stronger (earlier or later LSTs than the `foreground-quiet' range used in forming power spectra) --- doing so may yield a better description of the foregrounds that we desire to down-weight. Fundamentally, each of these examples are similar in that they rely on a computation of $\hat{\textbf{C}}$ from data that is similar but not exactly the same as the data that is being down-weighted. Ideally this would be effective in down-weighting shared contaminants yet avoid signal loss from over-fitting EoR modes in the power spectrum dataset itself.
